---
title: "Study of Severe Weather Events and their impact on population health and economic consequences"
author: "Venkata Yerubandi"
date: "January 23, 2015"
output: html_document
---
#### Synopsis
All weather events have health(injuries/fatalities) and economic consequences(property/crop damages) for counties. 

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database

[National Weather Service Storm Data Documentation ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

to address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

####  Data Processing 
```{r, echo=FALSE, cache=TRUE}
setwd("/Users/venkata/Downloads/coursera/Reproducible Research/Week 3/Assignment")
require("R.utils")
library(car)

library(ggplot2)
```

```{r, echo=TRUE, cache=TRUE}
# unzip the data file 
bunzip2("repdata-data-StormData.csv.bz2","repdata-data-StormData.csv",remove=FALSE,overwrite=TRUE)
stormdata<-read.csv("repdata-data-StormData.csv")

# look at the storm data topology
str(stormdata)

# look at stormdata$PROPDMGEXP and stormdata$CROPDMGEXP
# expenditure is encoded as B for Billion etc
unique(stormdata$PROPDMGEXP)
unique(stormdata$CROPDMGEXP)

# normalize expenditure indicators
# B is encoded as numeric billion etc
# assumption - a expenditore of a numeric 2 is encoded as 10^2 
# +/- are assigned a 0
stormdata$PROPDMGEXP <- factor(tolower(stormdata$PROPDMGEXP))
stormdata$CROPDMGEXP <- factor(tolower(stormdata$CROPDMGEXP))
stormdata$PROPDMGEXP <- as.numeric(recode(as.character(stormdata$PROPDMGEXP), 
                                          "'0'=1;'1'=10;'2'=10^2;'3'=10^3;'4'=10^4;'5'=10^5;'6'=10^6;'7'=10^7;'8'=10^8;'b'=10^9;'h'=10^2;'k'=10^3;'m'=10^6;'-'=0;'?'=0;'+'=0"))
stormdata$CROPDMGEXP <- as.numeric(recode(as.character(stormdata$CROPDMGEXP), 
                                          "'0'=1;'1'=10;'2'=10^2;'3'=10^3;'4'=10^4;'5'=10^5;'6'=10^6;'7'=10^7;'8'=10^8;'b'=10^9;'h'=10^2;'k'=10^3;'m'=10^6;'-'=0;'?'=0;'+'=0"))
# calculate property and crop damage values in dollars
stormdata$PROPDMGDOLLAR <- stormdata$PROPDMG * stormdata$PROPDMGEXP
stormdata$CROPDMGDOLLAR <- stormdata$CROPDMG * stormdata$CROPDMGEXP
```

```{r, echo=FALSE, cache=TRUE}
# multiplot function from  Winston Chang's R cookbook
# we use it to fit multiple plots in a single plot
multiplot <- function(..., plotlist=NULL, cols) {
    
    require(grid)
    
    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)
    
    numPlots = length(plots)
    
    # Make the panel
    plotCols = cols                          # Number of columns of plots
    plotRows = ceiling(numPlots/plotCols) # Number of rows needed, calculated from # of cols
    
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(plotRows, plotCols)))
    vplayout <- function(x, y)
        viewport(layout.pos.row = x, layout.pos.col = y)
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
        curRow = ceiling(i/plotCols)
        curCol = (i-1) %% plotCols + 1
        print(plots[[i]], vp = vplayout(curRow, curCol ))
        }
    
    }

```

##### Creating Aggregations (Injuries/Fatalities by event types , Crop/Property damages by event types)

For computing all aggregations we performing the following sequence of steps 

1. Aggregate the desired column across event type

2. Filter out the values which do not add value like rows with a zero value 

3. Sort the result and pick top 10 items

4. Use the top 10 items in plots 

Aggregrate injuries across event type
```{r, echo=TRUE, cache=TRUE}

# 1a. Aggregrate injuries across evttype
totalinjuriesbyevttype<-aggregate(stormdata$INJURIES, by=list(evtype=stormdata$EVTYPE), FUN=sum)
#rename columns 
colnames(totalinjuriesbyevttype) <- c("evtype","injuries")
# remove all the rows with 0 values ( i.e no affect on population health )
nonzerototalinjuriesbyevttype<-totalinjuriesbyevttype[totalinjuriesbyevttype$injuries!=0,]
#head(nonzerototalinjuriesbyevttype)
# Since there are a lot of nonzero vectors , sort and plot the 
# top 20
sortedtotalinjuriesbyevttype<-nonzerototalinjuriesbyevttype[order(-nonzerototalinjuriesbyevttype$injuries),]
top10sortedtotalinjuriesbyevttype<-sortedtotalinjuriesbyevttype[1:10,]
#head(top10sortedtotalinjuriesbyevttype)
```


 Aggregrate fatalities across event type
 
```{r, echo=TRUE, cache=TRUE}
# 1b. Aggregrate fatalities across evttype
totalfatalitiesbyevttype<-aggregate(stormdata$FATALITIES, by=list(evtype=stormdata$EVTYPE), FUN=sum)
#rename columns 
colnames(totalfatalitiesbyevttype) <- c("evtype","fatalities")
# remove all the rows with 0 values ( i.e no affect on population health )
nonzerototalfatalitiesbyevttype<-totalfatalitiesbyevttype[totalfatalitiesbyevttype$fatalities!=0,]
#head(nonzerototalfatalitiesbyevttype)
# Since there are a lot of nonzero vectors , sort and plot the 
# top 20
sortedtotalfatalitiesbyevttype<-nonzerototalfatalitiesbyevttype[order(-nonzerototalfatalitiesbyevttype$fatalities),]
top10sortedtotalfatalitiesbyevttype<-sortedtotalfatalitiesbyevttype[1:10,]
#head(top10sortedtotalfatalitiesbyevttype)
```

Aggregrate PROPDMGEXP across event type

```{r, echo=TRUE, cache=TRUE}
# 2a. Aggregrate PROPDMGEXP across evttype
totalpropdmgbyevttype<-aggregate(stormdata$PROPDMGDOLLAR, by=list(evtype=stormdata$EVTYPE), FUN=sum)
#rename columns 
colnames(totalpropdmgbyevttype) <- c("evtype","propdamage")
# remove all the rows with 0 values
nonzerototalpropdmgbyevttype<-totalpropdmgbyevttype[totalpropdmgbyevttype$propdamage!=0,]
#head(nonzerototalfatalitiesbyevttype)
# Since there are a lot of nonzero vectors , sort and plot the 
# top 20
sortedtotalpropdmgbyevttype<-nonzerototalpropdmgbyevttype[order(-nonzerototalpropdmgbyevttype$propdamage),]
top10sortedtotalpropdmgbyevttype<-sortedtotalpropdmgbyevttype[1:10,]
```

Aggregrate CROPDMGEXP across event type
```{r, echo=TRUE, cache=TRUE}
# 2b. Aggregrate CROPDMGEXP across evttype
totalcropdmgbyevttype<-aggregate(stormdata$CROPDMGDOLLAR, by=list(evtype=stormdata$EVTYPE), FUN=sum)
#rename columns 
colnames(totalcropdmgbyevttype) <- c("evtype","cropdamage")
# remove all the rows with 0 values 
nonzerototalcropdmgbyevttype<-totalcropdmgbyevttype[totalcropdmgbyevttype$cropdamage!=0,]
#head(nonzerototalfatalitiesbyevttype)
# Since there are a lot of nonzero vectors , sort and plot the 
# top 20
sortedtotalcropdmgbyevttype<-nonzerototalcropdmgbyevttype[order(-nonzerototalcropdmgbyevttype$cropdamage),]
top10sortedtotalcropdmgbyevttype<-sortedtotalcropdmgbyevttype[1:10,]
```


##### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

To answer the above question , we use sorted injuries and fatalities calculated across evttype and plot them. 

```{r, echo=TRUE, fig.width=40 , fig.height=15}

ipbet<-ggplot(top10sortedtotalinjuriesbyevttype, aes(x=evtype,y=injuries)) + geom_bar(aes(fill=evtype), stat="identity")+ggtitle("Injuries aggregated across event types")

fpbet<-ggplot(top10sortedtotalfatalitiesbyevttype, aes(x=evtype,y=fatalities)) + geom_bar(aes(fill=evtype), stat="identity")+ggtitle("Fatalities aggregated across event type")

multiplot(ipbet , fpbet , cols=2)
````

###### Results

Events (as indicated in the EVTYPE variable)  most harmful with respect to population health are 

Top injuries by event type 
```{r, echo=TRUE}
head(top10sortedtotalinjuriesbyevttype$evtype)
```

Top fatalities by event type 
```{r, echo=TRUE}
head(top10sortedtotalfatalitiesbyevttype$evtype)
```

##### Across the United States, which types of events have the greatest economic consequences?

To answer the above question , we use sorted crop damage  and property damage expenditures calculated across evttype and plot them. 

```{r, echo=TRUE, fig.width=40 , fig.height=15}
library(ggplot2)
propdamagebet<-ggplot(top10sortedtotalpropdmgbyevttype, aes(x=evtype,y=propdamage)) + geom_bar(aes(fill=evtype), stat="identity")+ggtitle("Property damage aggregated across event types")

cropdamagebet<-ggplot(top10sortedtotalcropdmgbyevttype, aes(x=evtype,y=cropdamage)) + geom_bar(aes(fill=evtype), stat="identity")+ggtitle("Crop damage aggregated across event type")

multiplot(propdamagebet , cropdamagebet , cols=2)
````

###### Results

Events with greatest economic consequences 

Top property damages by event type 
```{r, echo=TRUE}
head(top10sortedtotalpropdmgbyevttype$evtype)
```

Top crop damages by event type 
```{r, echo=TRUE}
head(top10sortedtotalcropdmgbyevttype$evtype)
```

